Measurement Bias from Address Aliasing

### 2016-01-09 paper.tex
Decrease listing font size
Reduce size of abstract

\lstset{language=C, basicstyle=\ttfamily, columns=flexible, belowskip=0pt}

@@

\author{\IEEEauthorblockN{Authors Name/s per 1st Affiliation (Author)}
\IEEEauthorblockA{line 1 (of Affiliation): dept. name of organization\\
line 2: name of organization, acronyms acceptable\\
line 3: City, Country\\
line 4: Email: name@xyz.com}
\and
\IEEEauthorblockN{Authors Name/s per 2nd Affiliation (Author)}
\IEEEauthorblockA{line 1 (of Affiliation): dept. name of organization\\
line 2: name of organization, acronyms acceptable\\
line 3: City, Country\\
line 4: Email: name@xyz.com}
}

@@

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
Understanding program behavior and obtaining accurate measurements is important in performance analysis.
However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.

This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
We demonstrate this by implementing runtime detection and correction of aliasing conditions.
For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in one of our examples.
Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
Our results can enable new architecture specific optimization strategies to account for this phenomenon.
\end{abstract}

@@

\subsection{Microkernel Analysis}
\label{sec:microkernel}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:microkernel}.
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
The following outlines our data and analysis of this program conducted on the ``Haswell'' configuration.
Similar experiments were conducted on the other experimental setups listed in Table~\ref{tab:machines} as well, all arriving at the same conclusion.

@@

%\newpage
\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the location of the stack at runtime is generally unknown.
In addition to being offset by environment variables, the initial stack address can also be perturbed by for example program arguments or address randomization.
Although one can not easily know if a collision is going to happen for a given environment a priori, it is possible to change the program to account for possible alias conditions.
A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.

@@

\subsection{Other Architectures}
While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
Figure~\ref{fig:microkernel-all-archs} shows additional cycle count plots for ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
On this architecture, the \emph{load\_blocks.overlap\_store} event is used to indicate address aliasing.
Notably, this event was also found to be significant in the original paper, where the authors used a similar setup~\cite{Mytkowicz:2009:WrongData}.
We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
Still, this only explains part of what is going on in Figure~\ref{fig:motivation}, as there seem to be yet another bias effect at every fourth environment increment.
These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
Our assumption is that this is a different phenomenon, not related to 4K aliasing.
We were only able to observe the effect on this particular architecture, and did not investigate the root cause further.

@@

Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between the heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
Numbers on the x axis represent the amount of offset, measured in number of \texttt{sizeof(float)} bytes.
An offset of zero is the default behavior for this program when using \texttt{malloc} and moderately large inputs.
Both for optimization level 2 and 3, the default alignment is close to worst case performance.
The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
This phenomenon is only observed for address separations close to zero, thus we only show the first 20 data points.
If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.

@@

Overall, it seems reasonable to conclude that the resource stalling is causing the slowdown, ultimately generated by false dependencies from address aliasing.
We do not attempt to pinpoint exactly which accesses generates these conflicts, but at a high level the CPU falsely assumes dependencies between \texttt{in[i]} and \texttt{out[i]}.
By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
This is a speedup on top of already aggressive compiler optimization.
% Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
% We do not attempt to pinpoint exactly which assembly instructions might alias, but one can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.

\subsection{Ways to Deal with Heap Address Aliasing}
Performance penalties caused by address aliasing can be significant, creating bias towards certain memory layouts.
However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted and to some extent eliminated in software.
The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
We identify some mitigation or optimization techniques that can be used:

\subsubsection{Mark buffers with \texttt{restrict}}
In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
This limits the extent generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
% \begin{lstlisting}[breaklines=true]
%     void conv(int n, const float * restrict in, float * restrict out);
% \end{lstlisting}
Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% on optimization level O2.
Alias events is reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results

@@

Cycles executed and alias events for each library is shown in Figure~\ref{fig:blas}.
Each library behaves differently, but in all cases we identify a significant amount of alias. 
The diagrams can be understood the same way as for the convolution example; partial address overlap occur when the address suffix delta of $A$ and $\boldsymbol{y}$ is close to zero.
Libblas generates the most distinct pattern, with a relatively clear spike for address deltas between 0x10 and 0x60.
% NB: Taken from analysis directory, update manually on new results
Correlation between cycles executed and address aliasing events were calculated to 95\% for libblas, 81\% for libatlas and 45\% for libopenblas.

@@

%\newpage
\section{Conclusions}
\label{sec:conclusions}
We have shown how address aliasing affects program performance under different memory contexts, and how it can explain certain cases of measurement bias.
The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
This phenomenon is likely present on most consumer hardware based on Intel platforms today, ranging from ``Core'' to ``Haswell'' microarchitectures.

In general, any change to how data is laid out in memory layout can potentially introduce bias effects from address aliasing.
Analyzing an example with bias to environment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
Dynamically allocated data is controlled by heap allocators, which we introduce as another source of bias.
Comparing four different libraries, we show that most implementations tend to favor page alignment for larger requests.
This means that structures such as vectors and matrices are likely to alias by default, which can be worst case scenarios for many algorithms.
%Address aliasing can have significant performance impact for algorithms operating on heap allocated buffers, which 
We analyze an example with up to $2x$ variation in cycle count between different heap alignments, showing that address aliasing can have a significant performance impact.
We are able to identify instances of alias issues in matrix-vector multiplication routines for different BLAS libraries, proving that this phenomenon also appear in real, performance critical code.

>>>>

\lstset{language=C, basicstyle=\ttfamily\small, columns=flexible, belowskip=0pt}

@@

\author{\IEEEauthorblockN{Lars Kirkholt Melhus, Rune Erlend Jensen}
\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
NTNU, Norway\\
lars.kirkholt.melhus@gmail.com, runeerle@idi.ntnu.no}
}

@@

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
%Understanding program behavior and obtaining accurate measurements is important in performance analysis.
%However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
%Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
In this paper, we identify \emph{address aliasing} as an important underlying mechanism causing measurement bias on modern Intel microarchitectures.
By analyzing hardware performance counters, we show how bias can occur from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
%Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
%We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.

%This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
We demonstrate ways to deal with this type of bias, through runtime detection and correction of aliasing conditions.
This is especially important for heap allocations, as most allocators tend to produce worst case alignment by default; favoring page alignment.
%For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in already optimized code.
%Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
Our results can enable new architecture specific optimization strategies to account for this phenomenon.
\end{abstract}

@@

\subsection{Microkernel Analysis}
\label{sec:microkernel}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:microkernel}.
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
The following data and analysis of this program is from the ``Haswell'' configuration.
Similar experiments were conducted on the other experimental setups listed in Table~\ref{tab:machines}, all arriving at the same conclusion.

@@

%\newpage
\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the location of the stack at runtime is generally unknown.
In addition to being offset by environment variables, the initial stack address can also be perturbed by for example program arguments or address randomization.
Although one can not easily know if a collision is going to happen for a given environment a priori, it is possible to change the program to account for possible alias conditions.
A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables further down the stack, and the alias condition is avoided.

@@

\subsection{Other Architectures}
While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
Figure~\ref{fig:microkernel-all-archs} shows additional cycle count plots for ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
On this architecture, the \emph{load\_blocks.overlap\_store} event is used to indicate address aliasing.
Notably, this event was also found to be significant in the original paper, where the authors used a similar setup~\cite{Mytkowicz:2009:WrongData}.
We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
Still, this only explains part of what is going on in Figure~\ref{fig:motivation}, as there seem to be yet another bias effect at every fourth environment increment.
These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
%Our assumption is that this is a different phenomenon, not related to 4K aliasing.
We were only able to observe the effect on this particular architecture, and did not investigate the root cause.% is related to 4K aliasing.

@@

Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between the heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
Numbers on the x axis represent the amount of offset, measured in number of \texttt{sizeof(float)} bytes.
An offset of zero is the default behavior for this program when using \texttt{malloc} and moderately large inputs.
Both for optimization level 2 and 3, the default alignment is close to worst case performance.
The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
This effect is only observed for address separations close to zero, thus we only show the first 20 data points.
If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.

@@

Overall, it seems reasonable to conclude that the resource stalling is causing the slowdown, ultimately generated by false dependencies from address aliasing.
We do not attempt to pinpoint exactly which accesses generates these conflicts, but at a high level the CPU falsely assumes dependencies between \texttt{in[i]} and \texttt{out[i]}.
By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
%This is a speedup on top of already aggressive compiler optimization.
% Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
% We do not attempt to pinpoint exactly which assembly instructions might alias, but one can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.

\subsection{Ways to Deal with Heap Address Aliasing}
Performance penalties caused by address aliasing can be significant, creating bias towards certain memory layouts.
However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted and to some extent eliminated in software.
The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
We identify some relevant optimization techniques:

\subsubsection{Mark buffers with \texttt{restrict}}
In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
This limits the extent generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
% \begin{lstlisting}[breaklines=true]
%     void conv(int n, const float * restrict in, float * restrict out);
% \end{lstlisting}
Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% and alias events by approximately 10 million on optimization level O2.
%Alias events are reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results

@@

Cycles executed and alias events for each library is shown in Figure~\ref{fig:blas}.
Each library behaves differently, but in all cases we identify a significant amount of alias. 
The diagrams can be understood the same way as for the convolution example; partial address overlap occur when the address suffix delta of $A$ and $\boldsymbol{y}$ is close to zero.
Libblas generates the most distinct pattern, with a clear spike for address deltas between 0x10 and 0x60.
% NB: Taken from analysis directory, update manually on new results
Correlation between cycles executed and address aliasing events were 95\% for libblas, 81\% for libatlas and 45\% for libopenblas.

@@

%\newpage
\section{Conclusions}
\label{sec:conclusions}
We have shown how address aliasing affects program performance under different memory contexts, and how it can explain certain cases of measurement bias.
%The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
Speculative and out-of-order memory operations handling only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
This effect is observed on Intel platforms, ranging from ``Core'' to ``Haswell'' microarchitectures.

In general, any change to how data is laid out in memory layout can potentially introduce bias effects from address aliasing.
Analyzing an example with bias to environment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
Dynamically allocated data is controlled by heap allocators, which we introduce as another source of bias.
Comparing four different libraries, we show that most implementations tend to favor page alignment for larger requests.
This means that structures such as vectors and matrices are likely to alias by default, which can be worst case scenarios for many algorithms.
%Address aliasing can have significant performance impact for algorithms operating on heap allocated buffers, which 
We analyze an example with up to $2x$ variation in cycle count between different heap alignments, showing that address aliasing can have a significant performance impact.
We are able to identify instances of alias issues in matrix-vector multiplication routines for different BLAS libraries, showing that this effect also appears in real, performance critical code.
